### 2017-10-10 response/response-jde-1.tex

To further investigate the role of abstinence we added a new
table to the Appendix, which further subdivides the results in Table 6.
Appendix Table A-8 shows, not surprisingly, that abstinence is the most 
effective traditional contraception, although only in combination 
with a crop loss as in the table in the main text.
A major caveat is that only 19 women report using abstinence only
across the three waves used for these regressions.
Modern contraceptive also are mainly effective in combination with
a crop loss, and the direct effect of using modern contraceptives
is not statistically significant.
We discuss these results toward the end of the section on 
``Conscious Decision or Unintended Consequences?''.

% Reduction in pregnancy is 0.3 for abstinence with crop loss
% Reduction in pregnancy is 0.2 for modern contraceptives with crop loss

\item The authors analyze the fertility of women who were married at the
start of the sample period. As a separate analysis, it would be
interesting to know whether marriage hazards change with crop loss in
the parents' home, although perhaps the authors do not have the sample
size to study this question.

[RESPONSE:] That is, indeed, a very interesting question, but we have 
decided not to pursue it at the moment for three reasons.
First, there are already two papers that examine this question, 
albeit for other countries \citep{mbiti08,Hoogeveen2011}.
Second, as those two papers show, the length of our paper would have
to increase substantially to do this question justice, and the Editor
has asked us to not increase the length of the paper.
Finally, the sample size is, indeed, on the small size to really get 
at this question.

@@

\item The survey measures crop loss continuously, and the indicator for
crop loss greater than 200 TZS is the authors' creation. They show in
the appendix that they obtain similar but less significant results if
they use the continuous measure (in levels or logs) or an indicator for
any crop loss. The 200 TZS seems arbitrary and possibly ex post, since
it leads to the most attractive results of the three sets of results
provided. It would be useful to more transparently show readers the
underlying variation and possible non-linearities, perhaps with a binned
version of the variable.

[RESPONSE:]

Our original discussion of this issue was scattered throughout the
original version of the paper.
We have now combined and expanded the discussion in the
Robustness section.
Before we respond, it is worth pointing out why we present the dummy
cut-off results in the paper.
We presented this paper a number of times before submission and whether 
we led with our dummy results or with the log results the discussant 
invariably asked us to focus on the other set of results.
We settled on the dummy results because they are easier to interpret 
for most people, but the downside is, as you point out, that the 
cut-off used is subject to discussion and that there is not one right level.

In many ways, the log specification in Appendix Table A-10 is more theoretically 
appealing because it captures that we should expect a given increase
in crop loss to have a diminishing impact on fertility related outcomes 
as the size of the crop loss increase (which is also why the linear version is 
not appealing; it assumes that a one TZS increase in crop loss has
the same effect whether it is at 0 or 50,000 TZS). 
Furthermore, the conclusion one would draw from the log results
is exactly the same as what we draw from the dummy approach.
Crop losses lead to statistically significant reductions in the likelihoods 
of pregnancy and birth, and statistically significant 
increases in the likelihood of using any contraceptives and using 
traditional contraceptives.
The differences from the dummy results are that the pregnancy and
birth results are slightly less statistically significant, whereas
the contraceptives results are slightly more statistically significant.
The main downside is that the log version is more difficult to interpret.

To show that the specific cut-off chosen does not have substantial effects on
the conclusions drawn, we have added Appendix Table A-9, which shows the main results
for different cut-offs together with the associated percentages of observations
that would be counted as having a crop loss.
Except for any crop loss (cut-off at zero TZS) the results
are consistent across all other specifications.
As the cut-off becomes very large statistically significance is lower
for pregnancy and births and eventually also for contraceptive use.
This is not surprising.
As we increase the cut-off there are more and more households 
that have experienced a crop loss, but are not counted as having a
crop loss, which will tend to bias results towards zero if there
is an effect of crop loss for levels lower than the cut-off 
(this is confirmed by regressions---not
shown---where we drop all observations with crop loss larger than zero but
smaller or equal to the cut-off and find substantially larger coefficients).

### 2017-10-10 response/response-jde-1.tex

\noindent Please find attached the revised version of our paper,
``Income Shocks, Contraceptive Use, and Timing of Fertility.''
We would like to thank the two referees for the very useful comments and
suggestions,
which we believe has improved the paper.

We rewrote the code for cleaning and analyzing the data from the ground up 
to make it easier to follow and eventually publish on GitHub.
This rewrite lowered the sample by three observations.
Including all married/partnered women up to age 45, rather than exclude
women younger than 18, the final sample size is, however, the same as 
before and the results are essentially unchanged.

To remain within the page limit we had to substantially reorganize
part of the paper.
The following is a list of the more substantial changes made:
\begin{enumerate}
\item We removed the map of Kagera.
\item In response to the reviewers' comments we have almost completely
rewritten the Robustness section as detailed below.
\item As part of the rewrite of the Robustness section we moved 
the discussion of age, wealth, and education results to a new Appendix
section since these results do little to clarify the mechanisms 
behind the main results.
We still address and incorporate the comments from reviewers that
relate to these outcomes.
\item We moved the discussion of estimation strategy for age,
wealth, and education estimation to the new Appendix section.
\item We have added another new Appendix section that examine results 
with including an additional lag for crop loss (see response to
Reviewer 1 below).
\item A number of new Appendix tables have been added in response
to various comments and other edited.
\end{enumerate}

We have tried to highlight the factors that can affect our identification 
strategy and reduce the use of causal language, but if we need to tone down 
the causality claims further we can do so.
We list below our responses to the individual comments and suggestions.

@@

To show that the specific cut-off chosen does not have substantial effects on
the conclusions drawn, we have added Appendix Table A-9, which shows the main results
for different cut-offs together with the associated percentages of observations
that would be counted as having a crop loss.
Except for any crop loss (cut-off at zero TZS) the results
are consistent across all other specifications.
As the cut-off becomes very large statistically significance is lower
for pregnancy and births and eventually also for contraceptive use.
This is not surprising.
As we increase the cut-off there are more and more households 
that have experienced a crop loss, but are not counted as having a
crop loss, which will tend to bias results towards zero if there
is an effect of crop loss for levels lower than the cut-off 
(this is confirmed by regressions---not
shown---where we drop all observations with crop loss larger than zero but
smaller or equal to the cut-off and find substantially larger coefficients).

\item The authors use causal language throughout the paper, describing
their results as effects. Of course, crop loss may be far from randomly
assigned, so this language is not appropriate.

[RESPONSE:]

We have reduce the use of causal language throughout the paper and
added this as an additional caveat to our results in the Conclusion.
Below we discuss factors that may explain the distribution of
crop loss.

### 2017-10-10 paper/incomeShocks-jde-r1.tex
Respond to R1 on causality

\begin{abstract} 
% JDE abstract
\noindent This paper examines the relationship between income 
shocks and fertility decisions. 
Using panel data from Tanzania, we estimate the impact of agricultural shocks 
on 
pregnancies, 
births, 
and 
contraception use.
The likelihood of 
pregnancies and childbirth are significantly lower for households that experience
a crop shock.
Furthermore, women significantly increase their contraception use in response 
to crop losses. 
The increase in contraceptive use comes almost entirely from traditional 
contraceptive methods, such as abstinence and the rhythm method.
We argue that these changes in behavior are the result of deliberate decisions of
the households rather than the shocks' effects on other factors that influence
fertility, such as women's health status, the absence or migration of a spouse, 
the dissolution of partnerships, or the number of hours worked. 
We also show that, although traditional contraceptives have low overall
efficacy, households with a strong incentive to postpone fertility
are very effective at using them.

@@

We use longitudinal data from the Kagera region in Tanzania with information on 
accidental crop loss, fertility, and contraceptive use to estimate individual 
level fixed effects models of the effects of income shocks on both timing of 
fertility and contraceptive use.
We show that income shocks decrease the likelihood of pregnancies and births.
At the same time, shocks significantly increase use of contraceptives.
This increase mainly occurs through the use of traditional contraceptives, 
which are the dominant form of contraception in the area.
We argue that the postponement of fertility in households experiencing income 
shocks is the result of the increased contraception use rather than a biological 
response or separation of spouses.

@@

Two main results stand out.
First, there are statistically significant and negative effects of crop loss on the
likelihood of both pregnancies and births.
Our preferred estimates---woman level fixed effects---show that experiencing 
crop loss decreases the likelihood of being pregnant by 10 percentage points 
and the likelihood of a birth by 17 percentage points.
These effects are substantial; on average across the four survey rounds, approximately 
17\% of women were pregnant and 17\% had given birth in the preceding period. 
Second, the absolute values of the community fixed effects point estimates are 
larger than the OLS estimates for both outcomes and the individual level
fixed effects estimates are larger than the community fixed effects estimates.%
\footnote{
Full results for the OLS and community fixed effects models are in shown in 
Table \ref{tab:ols_fertility}.
Table \ref{tab:com_cluster_birth} shows the results with clustering at community level
to account for potential spacial correlation in shocks.
Clustering at community level leads to slightly lower statistically significance 
for pregnancies and slightly higher statistically significance for births.
}
This suggests that unobservable characteristics, at both community
and individual level, correlate with a higher likelihood of experiencing 
a shock and lower likelihood of using contraceptives.

@@

The final question is how the effect on postponement of fertility 
is so strong when there is little use of modern contraceptives.
In Table \ref{tab:effectiveness}, we examine to what extent shocks cause
a delay in pregnancy through the increased use of contraceptives by
estimating the following: 
the effects of crop loss between the prior survey and the
survey before that, contraceptive use at the prior survey round, and their
interaction on whether the respondent is currently pregnant.
Contraceptive use is clearly endogenous, but this can provide 
suggestive evidence on whether shocks do delay
pregnancy through increased use of contraceptives.

@@

We estimate the relationship between household income shocks and decisions
on timing of fertility. 
To account for potential correlation between unobservable household
characteristics and both shocks and fertility decisions, we employ a fixed effect model.
Our estimates show that both the likelihood of pregnancy and birth decrease in
response to a crop loss and that contraceptive use, especially use of traditional 
methods, such as abstinence and rhythm, significantly increases.

We examine a number of causes that may explain the reduction 
in fertility following a crop loss.
Crop loss shocks do not worsen women's health enough to 
reduce pregnancies and childbirths to the extent that we find.
Furthermore, with no significant change in the likelihood of migration or 
dissolution of partnership in response to crop loss, there is no evidence 
that the reduction is due to physical separation of spouses or partner.
As expected, the number of hours worked increases for women after a shock, 
but the effect is unlikely to be large enough to affect fertility 
through a change in coital frequency.

Our main conclusion is that households consciously delay 
fertility through contraceptive use to cope with income shocks. 
This is especially of interest since this is accomplished 
predominately through the use of traditional contraceptives.
If traditional contraceptives have low efficacy, a natural question is 
how the substantial reductions in likelihood of births and pregnancy 
come about.
We present suggestive evidence that show that only those households
with a strong incentive, i.e. those exposed to crop loss, are 
effective users of traditional contraceptives.%
\footnote{
This is in line with the results in \cite{Rosenzweig1989},
where women with the highest opportunity cost of children
are also the most effective user of traditional contraceptives,
although the setting is very different.
}

There are two potential caveats to our results.
First, the sample size is small and the survey rounds
could have been further apart, but the longitudinal nature
of the data means that we can control for unobservable 
time-invariant individual and household characteristics.
Second, not surprisingly there appears to be measurement 
errors.
% Reported age and education levels are not always consistent
% across rounds, which is an indication that we probably also
% have measurement errors in the other variables, although it is
% harder to verify.
To the extent that these measurement errors are classical 
measurement errors, this biases our results towards zero,
making it harder to show an effect of crop loss.%
\footnote{
The problems with measurement error, combined with the small
sample size, may also explain why we do not find clear
differences in the response to crop loss across age groups,
education, and wealth levels.
}

>>>>

\begin{abstract} 
% JDE abstract
\noindent This paper examines the relationship between income 
shocks and fertility decisions. 
Using panel data from Tanzania, we estimate the impact of agricultural shocks 
on 
pregnancies, 
births, 
and 
contraception use.
The likelihood of 
pregnancies and childbirth are significantly lower for households that experience
a crop shock.
Furthermore, women have significantly higher contraception use after 
crop losses. 
The increase in contraceptive use comes almost entirely from traditional 
contraceptive methods, such as abstinence and the rhythm method.
We argue that these changes in behavior are the result of deliberate decisions of
the households rather than the shocks' effects on other factors that influence
fertility, such as women's health status, the absence or migration of a spouse, 
the dissolution of partnerships, or the number of hours worked. 
We also show that, although traditional contraceptives have low overall
efficacy, households with a strong incentive to postpone fertility
are very effective at using them.

@@

We use longitudinal data from the Kagera region in Tanzania with information on 
accidental crop loss, fertility, and contraceptive use to estimate individual 
level fixed effects models of the effects of income shocks on both timing of 
fertility and contraceptive use.
We show that income shocks are associated with a decrease the likelihood of 
pregnancies and births.
At the same time, use of contraceptives significantly increase after a shock.
This increase mainly occurs through the use of traditional contraceptives, 
which are the dominant form of contraception in the area.
We argue that the postponement of fertility in households experiencing income 
shocks is likely the result of the increased contraception use rather than a 
biological response or separation of spouses.

@@

Two main results stand out.
First, there are statistically significant and negative effects of crop loss on the
likelihood of both pregnancies and births.
Our preferred estimates---woman level fixed effects---show that experiencing 
crop loss is associated with a 10 percentage points decrease in the likelihood 
of being pregnant and a 17 percentage points reduction in the likelihood of a birth.
These estimates are substantial; on average across the four survey rounds, approximately 
17\% of women were pregnant and 17\% had given birth in the preceding period. 
Second, the absolute values of the community fixed effects point estimates are 
larger than the OLS estimates for both outcomes and the individual level
fixed effects estimates are larger than the community fixed effects estimates.%
\footnote{
Full results for the OLS and community fixed effects models are in shown in 
Table \ref{tab:ols_fertility}.
Table \ref{tab:com_cluster_birth} shows the results with clustering at community level
to account for potential spacial correlation in shocks.
Clustering at community level leads to slightly lower statistically significance 
for pregnancies and slightly higher statistically significance for births.
}
This suggests that unobservable characteristics, at both community
and individual level, correlate with a higher likelihood of experiencing 
a shock and lower likelihood of using contraceptives.

@@

The final question is how the effect on postponement of fertility 
is so strong when there is little use of modern contraceptives.
In Table \ref{tab:effectiveness}, we examine to what extent shocks are
associated with a delay in pregnancy through the increased use of 
contraceptives by estimating the following: 
the effects of crop loss between the prior survey and the
survey before that, contraceptive use at the prior survey round, and their
interaction on whether the respondent is currently pregnant.
Contraceptive use is clearly endogenous, but this can provide 
suggestive evidence on whether shocks do delay
pregnancy through increased use of contraceptives.

@@

We estimate the relationship between household income shocks and decisions
on timing of fertility. 
To account for potential correlation between unobservable household
characteristics and both shocks and fertility decisions, we employ a fixed effect model.
Our estimates show that both the likelihood of pregnancy and birth decrease 
after a crop loss and that contraceptive use, especially use of traditional 
methods, such as abstinence and rhythm, significantly increases.

We examine a number of potential explanations for the reduction 
in fertility following a crop loss.
Crop loss shocks do not worsen women's health enough to 
reduce pregnancies and childbirths to the extent that we find.
Furthermore, with no significant change in the likelihood of migration or 
dissolution of partnership in response to crop loss, there is no evidence 
that the reduction is due to physical separation of spouses or partner.
As expected, the number of hours worked increases for women after a shock, 
but the effect is unlikely to be large enough to affect fertility 
through a change in coital frequency.

Although we cannot entirely rule out that other, unobserved
factors play a role, our main conclusion is that households appear
to consciously delay fertility through contraceptive use to cope 
with income shocks. 
This is especially of interest since this is accomplished 
predominately through the use of traditional contraceptives.
If traditional contraceptives have low efficacy, a natural question is 
how the substantial reductions in likelihood of births and pregnancy 
come about.
We present suggestive evidence that show that only those households
with a strong incentive, i.e. those exposed to crop loss, are 
effective users of traditional contraceptives.%
\footnote{
This is in line with the results in \cite{Rosenzweig1989},
where women with the highest opportunity cost of children
are also the most effective user of traditional contraceptives,
although the setting is very different.
}

There are three potential caveats to our results.
First, the sample size is small and the survey rounds
could have been further apart, but the longitudinal nature
of the data means that we can control for unobservable 
time-invariant individual and household characteristics.
Second, not surprisingly there appears to be measurement 
errors.
% Reported age and education levels are not always consistent
% across rounds, which is an indication that we probably also
% have measurement errors in the other variables, although it is
% harder to verify.
To the extent that these measurement errors are classical 
measurement errors, this biases our results towards zero,
making it harder to show an effect of crop loss.%
\footnote{
The problems with measurement error, combined with the small
sample size, may also explain why we do not find clear
differences in the response to crop loss across age groups,
education, and wealth levels.
}
Third, although we cover a number of possible explanations
for what determines crop loss, it is possible that there may 
be other unobserved, time-varying household characteristics 
that influence whether a household experience a crop loss.
If that is the case the causality from crop to fertility decisions
may not hold.
